One-armed spirals in locally isothermal, radially structured self-gravitating discs

### 2015-01-09 paper/appendix.tex
very rough first revised version

If we associate the external potential with angular momentum 
$J_\mathrm{ext}  = M_\mathrm{ring}\overline{R}^2\Omega_p$, we can
calculate a growth rate $\gamma_L=\Gamma_L/J_\mathrm{ext}$. Then 
\begin{align}
  \frac{\gamma_L}{\Omega_p} = &\frac{\pi
    h}{3Q_L}\left(\frac{M_p}{M_*}\right)\left(\frac{R_L}{\overline{R}}\right)\left(\frac{R_c}{\overline{R}}\right) 
  ^3\left(\frac{R_L}{R_c}\right)^{-3/2}\notag\\ 
  &\times
  \left\{\frac{R_L}{\overline{R}}\left.\frac{db_{1/2}^1}{d\beta}\right|_L
    + 2\left[1 - 
      \left(\frac{R_c}{R_L}\right)^{-3/2}\right]b_{1/2}^1(\beta_L)\right\}^2. 
\end{align}
%\begin{align}
 % \frac{\gamma_L}{\Omega_p} \sim &\frac{\pi h}{3
  %  Q_L}\left(\frac{M_\mathrm{ring}}{M_*}\right)\left(\frac{R_L}{\overline{R}}\right)^{-1/2}\left(\frac{R_c}{\overline{R}}\right)^{5/2}\notag\\
 % &\times \left\{ \frac{R_L}{R_L - \overline{R}} - 2\left[1 -
  %    \left(\frac{R_c}{R_L}\right)^{-3/2}\right]\right\}^2\frac{R_c^2}{\left(R_L
   % - \overline{R}\right)^2}. 
%\end{align}
Inserting $h=0.05$, $Q_L=10$, $M_\mathrm{ring} = 0.05M_*$,
$R_L=7.2R_0$, $R_c=4.4R_0$ and $\overline{R}=1.5R_0$ from our fiducial
FARGO simulation, we get
\begin{align}
  \gamma_L \sim 5\times10^{-4}\Omega_p. 
\end{align}
%\begin{align}
 % \gamma_L \sim 0.012\Omega_p. 
%\end{align}
%This represents an upper limit on the effect of the OLR on the spiral
%disturbance, and is much smaller than the growth rate measured in the
%simulation ($\gamma\sim0.1\Omega_p$). We conclude that for our models,
%the OLR has negligible effect on the one-armed spiral.  
For the co-rotation torque, we use the result
\begin{align}
  \Gamma_c = \left.
    \pi^2\Phi_\mathrm{ext}^2\left(\frac{d\Omega}{dR}\right)^{-1}\frac{d}{dR}\left(\frac{2\Sigma}{\Omega}\right)\right|_{c}     
\end{align}
for a Keplerian disc, where subscript $c$ denotes evaluation at
co-rotation radius $R=R_c$. For a power-law surface density profile
$\Sigma\propto R^{-s}$ we have
\begin{align}
\frac{\gamma_c}{\Omega_p} = \frac{4}{3}\frac{\pi h}{Q_c}
\left(\frac{M_\mathrm{ring}}{M_*}\right)\left(\frac{R_c}{\overline{R}}\right)^4\left(s-\frac{3}{2}\right) 
\left[b_{1/2}^1(\beta_c)\right]^2    
\end{align}
%\begin{align}
%  \frac{\Gamma_C}{J_\mathrm{ext}\Omega_p} = \frac{4}{3}\frac{\pi h}{Q_C} \left(s -
 %   \frac{3}{2}\right)\left(\frac{M_\mathrm{ring}}{M_*}\right)\frac{R_c^4}{\overline{R}^2\left(R_c
 %     - \overline{R}\right)^2}.   
%\end{align}
Using the above parameter values with $s=2$ and $Q_c=10$, we obtain a
rate
\begin{align}
  \gamma_c\sim 6\times 10^{-4}\Omega_p. 
\end{align}

>>>>

If we associate the external potential with angular momentum 
$J_\mathrm{ext}  = M_\mathrm{ring}\overline{R}^2\Omega_p$, we can
calculate a growth rate $\gamma_L=\Gamma_L/J_\mathrm{ext}$. Then 
\begin{align}
  \frac{\gamma_L}{\Omega_p} = &\frac{\pi
    h}{3Q_L}\left(\frac{M_p}{M_*}\right)\left(\frac{R_L}{\overline{R}}\right)\left(\frac{R_c}{\overline{R}}\right) 
  ^3\left(\frac{R_L}{R_c}\right)^{-3/2}\notag\\ 
  &\times
  \left\{\frac{R_L}{\overline{R}}\left.\frac{db_{1/2}^1}{d\beta}\right|_L
    + 2\left[1 - 
      \left(\frac{R_c}{R_L}\right)^{-3/2}\right]b_{1/2}^1(\beta_L)\right\}^2. 
\end{align}
%\begin{align}
 % \frac{\gamma_L}{\Omega_p} \sim &\frac{\pi h}{3
  %  Q_L}\left(\frac{M_\mathrm{ring}}{M_*}\right)\left(\frac{R_L}{\overline{R}}\right)^{-1/2}\left(\frac{R_c}{\overline{R}}\right)^{5/2}\notag\\
 % &\times \left\{ \frac{R_L}{R_L - \overline{R}} - 2\left[1 -
  %    \left(\frac{R_c}{R_L}\right)^{-3/2}\right]\right\}^2\frac{R_c^2}{\left(R_L
   % - \overline{R}\right)^2}. 
%\end{align}
Inserting $h=0.05$, $Q_L=10$, $M_\mathrm{ring} = 0.05M_*$,
$R_L=7.2R_0$, $R_c=4.4R_0$ and $\overline{R}=1.5R_0$ from our fiducial
FARGO simulation, we get
\begin{align}
  \gamma_L \sim 5\times10^{-4}\Omega_p. 
\end{align}
%\begin{align}
 % \gamma_L \sim 0.012\Omega_p. 
%\end{align}
%This represents an upper limit on the effect of the OLR on the spiral
%disturbance, and is much smaller than the growth rate measured in the
%simulation ($\gamma\sim0.1\Omega_p$). We conclude that for our models,
%the OLR has negligible effect on the one-armed spiral.  
For the co-rotation torque, we use the result
\begin{align}
  \Gamma_c = \left.
    \pi^2\Phi_\mathrm{ext}^2\left(\frac{d\Omega}{dR}\right)^{-1}\frac{d}{dR}\left(\frac{2\Sigma}{\Omega}\right)\right|_{c}     
\end{align}
for a Keplerian disc, where subscript $c$ denotes evaluation at
co-rotation radius $R=R_c$. For a power-law surface density profile
$\Sigma\propto R^{-s}$ we have
\begin{align}
\frac{\gamma_c}{\Omega_p} = \frac{4}{3}\frac{\pi h}{Q_c}
\left(\frac{M_\mathrm{ring}}{M_*}\right)\left(\frac{R_c}{\overline{R}}\right)^4\left(s-\frac{3}{2}\right) 
\left[b_{1/2}^1(\beta_c)\right]^2    
\end{align}
%\begin{align}
%  \frac{\Gamma_C}{J_\mathrm{ext}\Omega_p} = \frac{4}{3}\frac{\pi h}{Q_C} \left(s -
 %   \frac{3}{2}\right)\left(\frac{M_\mathrm{ring}}{M_*}\right)\frac{R_c^4}{\overline{R}^2\left(R_c
 %     - \overline{R}\right)^2}.   
%\end{align}
Using the above parameter values with $s=2$ and $Q_c=10$, we obtain a
rate
\begin{align}
  \gamma_c\sim 6\times 10^{-4}\Omega_p. 
\end{align}
These estimates are much smaller than that due to the
imposed temperature gradient as measured in the simulations
($\gamma\sim∼0.1\Omega_p$). We conclude that for our disc models, 
the Lindblad and co-rotation resonances have negligible effects
on the growth of the one-armed spiral in the inner disc.

The torque exerted on the disc at the OLR by an external potential is
positive, while that at co-rotation depends 
on the gradient of potential vorticity there \citep{goldreich79}. For
our disc models with surface density $\Sigma\propto R^{-2}$ in the
outer disc, this co-rotation torque is positive. 
%Interpreting the one-armed spiral as an external po-
%tential for the outer disc, 
This means that the one-armed spiral loses
angular momentum by launching density waves with positive angular
momentum at the OLR, and by applying a 
positive co-rotation torque on the disc. In principle, this
interaction is destabilising because the one-armed spiral has 
negative angular momentum \citep{lin11b}. This
effect is small in our simulations, but it could be significant
in other parameter regimes.

### 2015-01-09 paper/discussion.tex
very rough first revised version

@@

  \subsubsection{Radial disc structure}
  In our simulations the 
  $m=1$ spiral is confined between two $Q$-barriers, where real solutions to the local
  dispersion relation is possible (Eq. \ref{wavenumber}). The
  existence of such a cavity is a result of the adopted surface
  density bump (Eq. \ref{sig_bump}).  
  Thus, in our disc models the main role of disc structure 
  and self-gravity is to allow a local $m=1$ mode to be set up, which is then 
  destabilised by the background torque.

@@

  %in fall/mass loading? 
  The surface density bump adopted in our numerical simulations might
  represent, for example, mass built-up in a 
  `dead zone' in protoplanetary 
  discs, where the magneto-rotational 
  instability is ineffective for angular momentum transport
  \citep{gammie96,turner08,landry13} and therefore has a reduced mass
  accretion rate. The dead zone becomes 
  self-gravitating with sufficient mass built-up 
  \citep{armitage01,martin12,martin12b,zhu09,zhu10,zhu10b,bae13}.

@@

  %particularly simple functional representation of the bump. a
  %realistic bump may be different

  Another possibility is a gap opened by a giant planet. In that case $Q$
  rises rapidly towards the gap centre since it is a region of low
  surface density, which can satisfy Eq. \ref{qb_cond}. Then the inner
  edge of our bump
  function mimics the outer gap edge. The outer gap edge is then a
  possible site for the growth of a low-frequency one-armed spiral
  through the background torque. 
 %or gap opened by a fragment

}

@@

{\bf
  \subsection{Motion of the central star} 
  For the simulations presented in this paper, the indirect potential
associated with the non-inertial reference frame is purposefully 
neglected to avoid complications arising from the motion of the
central star. For the adopted disc masses $M_d\lesssim 0.1 M_*$, this
effect is not expected to be significant
\citep{adams89,shu90,michael10}. Indeed, simulations including the indirect
potential, carried out in the early stages of this project, produced
similar results.

  \subsection{Role of Lindblad and co-rotation torques}
  One possible effect of self-gravity is to allow the one-armed spiral  
  to act as an external potential for the exterior disc. This is
  analogous to disc-satellite interaction 
  \citep{goldreich79}, where the embedded satellite exerts a torque on
  the disc at

In Appendix \ref{disc-planet} we estimate the magnitude of this effect
  using basic results from disc-planet theory \citep[see, e.g.][and
  references therein]{papaloizou07}.

}

>>>>

{\bf
  \subsection{Motion of the central star} 
  In our models we have purposefully neglected the indirect potential
  associated with a non-inertial reference frame to avoid
  complications arising from the motion of the central star.  
  It has been established that such motion, in response to an $m=1$
  disturbance in the disc, can render the disturbance
  unstable in a barotropic disc (so the background torque is
  ineffective).

  However, for the disc masses in our models, $M_d\lesssim 
  0.1 M_*$, this effect is not expected to be significant
  \citep{adams89,shu90,michael10}. Indeed, simulations including the
  indirect potential, carried out in the early stages of this project,
  produced similar results.

  \subsection{Role of Lindblad and co-rotation torques}
  One effect of self-gravity is to allow the one-armed spiral,
  confined to $R\sim R_0$, to act as an external potential for the
  exterior disc in $R\gtrsim R_0$. This is 
  analogous to disc-satellite interaction 
  \citep{goldreich79}, where the embedded satellite exerts a torque on
  the disc at Lindblad and co-rotation resonances.

  In Appendix \ref{disc-planet} we estimate the magnitude of this effect
  using basic results from disc-planet theory \citep[see, e.g.][and
  references therein]{papaloizou07}. There, we find that the angular
  momentum exchange between the one-armed spiral and the exterior disc
  is insignificant compared to the background torque.

  We confirmed this with additional FARGO simulations that exclude the
  co-rotation and outer Lindblad resonances (OLR) by reducing radial domain
  size, setting $R_\mathrm{max} = 3 R_0$. This model still developed
  the one-armed spiral. 
}

@@

  \subsubsection{Radial disc structure}
  In our simulations the 
  $m=1$ spiral is confined between two $Q$-barriers, where real solutions to the local
  dispersion relation is possible (Eq. \ref{wavenumber}). The
  existence of such a cavity is a result of the initial surface
  density bump (Eq. \ref{sig_bump}).  
  Thus, in our disc models the main role of disc structure 
  and self-gravity is to allow a local $m=1$ mode to be set up, which is then 
  destabilised by the background torque.

@@

  A surface density bump can develop in 
  `dead zones' of protoplanetary 
  discs, where there is reduced mass accretion because the magneto-rotational 
  instability is ineffective for angular momentum transport 
  \citep{gammie96,turner08,landry13}. The dead zone becomes 
  self-gravitating with sufficient mass built-up 
  \citep{armitage01,martin12,martin12b,zhu09,zhu10,zhu10b,bae13}.

  However, conditions in a dead zone may not be suitable for 
  sustaining a background torque because it may not cool/heat fast enough
  to maintain a fixed temperature profile. Recently, 
  \cite{bae14} presented sophisticated numerical models of
  dead zones, including a range of heating and 
  cooling processes, which show that dead zones developed large-scale
  (genuine) gravitational instabilities with multiple spiral
  arms. Although this does not prove absence of the background torque,
  it is probably insignificant compared to gravitational torques.

@@

  Another possibility is a gap opened by an embedded planet. In that case $Q$
  rises rapidly towards the gap centre since it is a region of low
  surface density. This can satisfy Eq. \ref{qb_cond}. Then the inner
  edge of our bump function mimics the outer gap edge. The outer gap
  edge is then a potential site for the growth of a low-frequency
  one-armed spiral through the background torque. However, the
  locally isothermal requirement limits this process to the
  outer disc.   
}

@@

### 2015-01-09 paper/paper.tex
updating...
very rough first revised version

\maketitle
\begin{abstract} 
  {\bf We describe a new mechanism that leads to the destabilisation of
 non-axisymmetric waves in astrophysical discs with 
  an imposed radial temperature gradient.  We use linear density wave
  theory 
  to show that non-axisymmetric perturbations generally do not conserve
  their angular momentum in the presence of a forced temperature
  gradient. This implies an exchange of angular momentum
  between linear perturbations and the background disc. In particular, 
  when the disturbance is a low-frequency
  trailing wave and the disc temperature decreases outwards, this 
  interaction is unstable and leads to the growth of the wave.  
  We demonstrate this phenomenon through numerical hydrodynamic
  simulations of locally isothermal discs in 2D using the FARGO
  code and in 3D with  the ZEUS-MP and PLUTO codes.  We consider
  radially structured discs with a self-gravitating sub-portion which remains
  stable in the absence of a temperature gradient. However,
  when a temperature gradient is imposed we observe exponential growth of an one-armed  
  spiral mode (azimuthal wavenumber $m=1$) with co-rotation radius 
  outside the bulk of the spiral arm, resulting in a nearly-stationary
  one-armed spiral pattern.  The development of this one-armed 
 spiral does not require the movement of the central star, as found in previous studies. 
 Because destabilisation by the imposed temperature gradient does 
 not explicitly require disc self-gravity, we suggest this mechanism may also
 affect low-frequency one-armed oscillations in non-self-gravitating
 discs. 
}
\end{abstract}

@@

% but also natural in GI.
However, spiral arms are also associated with global gravitational 
instability (GI) of differentially rotating discs
\citep{goldreich65,laughlin96b,laughlin98,nelson98,lodato05,forgan11}. Large-scale
spiral arms can provide significant angular momentum transport
necessary for mass accretion \citep{lynden-bell72, papaloizou91,balbus99,lodato04}, and
spiral structures due to GI are potentially observable with the Atacama Large 
Millimeter/sub-millimeter Array \citep{cossins10,dipierror14}.

% when/where does it happen? formation stage/dead zones (special for ppd)
GI can be expected in the earliest stages of 
circumstellar disc formation
\citep{kratter10b,inutsuka10,tsukamoto13}, and may be possible in the
outer parts of the disc \citep{rafikov05,matzner05,kimura12}.

@@

%brief history of m=1 
%stellar wobble from indiana group  
Single-arm spirals, or eccentric modes, corresponding to perturbations 
with azimuthal wavenumber $m=1$, have received interest in
the context of protoplanetary discs because of their global nature
\citep{adams89,heemskerk92,laughlin96,tremaine01,papaloizou02,hopkins10}. 
In the `SLING' mechanism proposed by \cite{shu90}, the $m=1$ spiral
instability is attributed to motion of the central star induced by the
one-armed perturbation and requires a massive disc \citep[the former may
have observable consequences, ][]{michael10}. However, our numerical
models do not have these features.

%these are self-gravitating instabilities

The aforementioned analytical studies have adopted barotropic, smooth disc models. 
In this work, we identify a mechanism to generate one-armed
spiral patterns that is related to a non-barotropic equation of
state and disc structure. We find that if the disc temperature or
sound-speed is a prescribed function of position, then 
one-armed spiral modes can grow inside a cavity that results from the
adopted disc structure.

Our study is based on direct hydrodynamic simulations using three
different grid-based codes, both in two-dimensions (2D) and
three-dimensions (3D). We also apply linear
theory to interpret simulation results. We find that, when the disc
temperature is prescribed, the usual statement for the conservation of
angular momentum for linear perturbations acquires a source term
proportional to the temperature gradient, which may act to destabilise
$m=1$ perturbations. Although we work with self-gravitating discs,
this source term does not depend explicitly on disc self-gravity. Thus, the
destabilisation effect we describe should also be applicable to
non-self-gravitating discs.

This paper is organised as follows. In \S\ref{model} we list the
governing equations and describe the disc equilibria under
consideration. We briefly review basic results from linear theory in
\S\ref{wkb} to aid our interpretation of numerical simulations. 
\S\ref{methods} describes the hydrodynamic codes
employed in this work and several diagnostic measures. Our main results are presented in 
\S\ref{results2d} for 2D simulations and \S\ref{results3d} for 3D
simulations, and we further analyse them in \S\ref{discussions}. 
We summarise in \S\ref{summary}, with a discussion on
speculations for future work.

%\section*{Acknowledgments}

\bibliographystyle{mn2e}
\bibliography{ref}

>>>>

\maketitle
\begin{abstract} 
  {\bf We describe a new mechanism that leads to the destabilisation of
 non-axisymmetric waves in astrophysical discs with 
 an imposed radial temperature gradient. This might apply, for
 example, to the outer parts of protoplanetary discs. We use linear
 density wave theory 
 to show that non-axisymmetric perturbations generally do not conserve
 their angular momentum in the presence of a forced temperature
 gradient. This implies an exchange of angular momentum
 between linear perturbations and the background disc. In particular, 
 when the disturbance is a low-frequency
 trailing wave and the disc temperature decreases outwards, this 
 interaction is unstable and leads to the growth of the wave.  
 We demonstrate this phenomenon through numerical hydrodynamic
 simulations of locally isothermal discs in 2D using the FARGO
 code and in 3D with  the ZEUS-MP and PLUTO codes.  We consider
 radially structured discs with a self-gravitating region which remains
 stable in the absence of a temperature gradient. However,
 when a temperature gradient is imposed we observe exponential growth of an one-armed  
 spiral mode (azimuthal wavenumber $m=1$) with co-rotation radius 
 outside the bulk of the spiral arm, resulting in a nearly-stationary
 one-armed spiral pattern.  The development of this one-armed 
 spiral does not require the movement of the central star, as found in previous studies. 
 Because destabilisation by the imposed temperature gradient does 
 not explicitly require disc self-gravity, we suggest this mechanism may also
 affect low-frequency one-armed oscillations in non-self-gravitating
 discs. 
}
\end{abstract}

@@

% but also natural in GI.
Spiral arms are also {\bf characteristic of} global gravitational 
instability (GI) {\bf in} differentially rotating discs
\citep{goldreich65,laughlin96b,laughlin98,nelson98,lodato05,forgan11}. Large-scale 
spiral arms can provide significant angular momentum transport
necessary for mass accretion \citep{lynden-bell72,
  papaloizou91,balbus99,lodato04}, and spiral structures due to GI are
potentially observable with the Atacama Large
Millimeter/sub-millimeter Array \citep{cossins10,dipierror14}. GI can
be expected in the earliest stages of circumstellar disc formation 
\citep{kratter10b,inutsuka10,tsukamoto13}, and may be possible in the
outer parts of the disc \citep{rafikov05,matzner05,kimura12}.

Single-arm spirals, or eccentric modes, corresponding to perturbations 
with azimuthal wavenumber $m=1$, have received interest in
the context of protoplanetary discs because of their global nature
\citep{adams89,heemskerk92,laughlin96,tremaine01,papaloizou02,hopkins10}. 
In the `SLING' mechanism proposed by \cite{shu90}, the $m=1$ {\bf
  gravitational spiral} instability is attributed to motion of the
central star induced by the one-armed perturbation and requires a
massive disc \citep[the former may have observable consequences,
][]{michael10}.

% The aforementioned analytical studies have adopted
% barotropic, smooth disc models.
% In this work, we identify a mechanism to generate one-
% armed spiral patterns that is related to a non-barotropic
% equation of state and disc structure.
% We find that if the disc temperature or sound-speed is a
% prescribed function of position, then one-armed spiral modes
% can grow inside a cavity that results from the adopted disc
% structure.

%The aforementioned analytical studies have adopted barotropic, smooth
%disc models.

{\bf
  In this work we identify a new mechanism that leads to the growth of
  one-armed spirals in astrophysical discs. We show
  that when the disc temperature is prescribed (called locally isothermal
  discs), the usual statement for the conservation of angular momentum
  for linear perturbations acquires a source term proportional to the
  temperature gradient. This permits angular momenum exchange between
  linear perturbations and the background disc. This `background
  torque' may act to destabilise low-frequency non-axisymmetric
  trailing waves when the disc temperature decrease outwards.       
}

{\bf We employ direct hydrodynamic simulations using three different
  grid-based codes to demonstrate how this `background torque' 
  can lead to the growth of one-armed spirals in radially structured, 
  self-gravitating discs. This is despite the fact that our disc
  models do not meet the requirements for the `SLING'
  mechanism. Although our numerical simulations consider
  self-gravitating discs, this `background torque' is generic for
  locally isothermal discs and its existence does not depend
  explicitly on disc self-gravity. Thus, the destabilisation effect we
  describe should also be applicable to non-self-gravitating discs.  
}

This paper is organised as follows. We describe the system of interest
and list the governing equations in 
\S\ref{model}. {\bf In \S\ref{wkb} we use linear theory  to show
  how a fixed temperature profile 
  can destabilise non-axisymmetric waves in discs.} \S\ref{methods} describes
the {\bf numerical setup and} hydrodynamic codes {\bf we use to 
  demonstrate the growth of one-armed spirals due to an imposed
  temperature gradient.} Our {\bf
  simulation} results are presented in  \S\ref{results2d} for two-
dimensional (2D) {\bf
  discs} and \S\ref{results3d} for three-dimensional (3D) {\bf discs}, and we further
discuss them in \S\ref{discussions}. We summarise in \S\ref{summary} 
with a discussion on speculations for future work.

{\bf
\section*{Acknowledgments}
This material is based upon work supported by the National Science Foundation under Grant No. 1228509.
}
\bibliographystyle{mn2e}
\bibliography{ref}

@@

%brief history of m=1 
%stellar wobble from indiana group

### 2015-01-09 paper/summary.tex
major revision in response to report
very rough first revised version

\section{Summary and conclusions}\label{summary}
In this paper, we have carried out direct numerical hydrodynamic
simulations of radially structured, self-gravitating locally
isothermal discs.     
We find such systems can be unstable to low-frequency perturbations
with azimuthal wavenumber $m=1$, which leads to the development of an one-armed 
trailing spiral that persist for at least $O(10^2)$ orbits. The spiral
pattern speed is smaller than the local disc rotation and
growth rates are $O(10^{-2}\Omega)$ which gives a characteristic
growth time of $O(10)$ orbits.   
%While this is slow
%compared to the local rotation, it is dynamical at co-rotation.

We used three independent numerical codes --- FARGO in 2D, ZEUS-MP and
PLUTO in 3D --- to show that the growth of  
one-armed spirals in our disc models is due to the imposed temperature 
profile and disc structure. Growth rates increased 
with the magnitude of the imposed temperature gradient, and one-armed
spirals did not develop in strictly isothermal simulations. The spiral
is confined between two $Q$-barriers owing to a surface density bump. 
We find the instability behaves similarly in 2D and 3D. However, in 3D
the spiral disturbance can become radially global away from 
the midplane.

We applied angular momentum conservation within linear theory to
explain our numerical results. We find a   
forced temperature gradient introduces a torque on linear 
perturbations. We called this the background torque because it
represents an exchange of angular momentum between the background disc
and the perturbations. This offers a previously unexplored pathway to
instability.

In the local approximation, we showed that this background torque is
negative for trailing spirals in discs with a fixed temperature or
sound-speed profile that decrease outwards. This torque enforces  
low-frequency $m=1$ modes because they are associated with negative
angular momentum. We therefore interpret the one-armed spiral instability
observed in our simulations as an initially neutral, tightly-wound
$m=1$ mode being destabilised by the imposed temperature gradient.

\subsection{Speculations and future work}
%numerical simulations to include cooling/heating
A locally isothermal equation of state represents the limit of
infinitely short cooling (and heating) timescales, so the disc
temperature instantly returns to its initial value when perturbed.  
This assumption can be relaxed by including an energy equation with
thermal relaxation. Preliminary FARGO simulations
indicate a thermal relaxation timescale $t_c < 0.1\Omega_k^{-1}$ is
needed for the one-armed spiral to develop. However, this value may
depend on disc parameters, and will be explored in a follow-up  
study.

% prelim sims show they are long lasting 
% balancing heating and cooling 
% should not get fragmentation (CR away from spiral) 
In the deeply non-linear regime, the one-armed spiral may 
shock and deposit negative angular momentum onto
the background disc. The spiral amplitude would saturate by gaining
positive angular momentum. However, if the temperature gradient is
maintained, it may be possible to achieve a balance between the gain
of negative angular momentum through the background torque, and the
gain of positive angular momentum through shock dissipation. We remark  
that fragmentation is unlikely because the co-rotation radius 
is outside the bulk of the spiral arm, in the non-self-gravitating
portion of the disc \citep{durisen08}. In order
to study these possibilities, improved numerical models are needed to
ensure total angular momentum conservation on timescales much longer
than that considered in this paper.

@@

In our models and interpretation, self-gravity coupled with the disc 
structure permits a neutral one-armed mode, which is then
destabilised by the temperature gradient. Other origins of the neutral
one-armed spiral should be investigated. One possibility is in Be star discs
\citep{rivinius13}, for which 
one-armed oscillations may explain long-timescale variations in their
emission lines \citep[see e.g.][and references
therein]{okasaki97,papaloizou06c,ogilvie08}. These studies employ
strictly isothermal disc models, so it would be interesting to explore
the effect of a temperature gradient on the stability of these
oscillations.

%vertical shear instability?

%linear theory: explicit calculations 
%For a self-consistent description of the instability, one needs to
%explicitly solve for linear stability problem.

%theoretical investigation on effect on other m modes

>>>>

 % Strictly speaking, the growth of the one-armed spiral
 %   observed in our simulations is \emph{not} a gravitational instability
 %   because destabilisation is through the forced temperature gradient
 %   and not disc self-gravity.  Nevertheless, we discuss below several issues 
 %   related to self-gravitating discs.

\section{Summary and conclusions}\label{summary}
{\bf
  In this paper, we have described a destabilising  
  effect of adopting a fixed temperature profile to model   
  astrophysical discs. By applying angular momentum conservation
  within linear theory, we showed that a forced temperature gradient 
  introduces a torque on linear perturbations. We call this the
  background torque because it represents an exchange of angular
  momentum between the background disc and the perturbations. This 
  offers a previously unexplored pathway to instability in locally
  isothermal discs.  
}

In the local approximation, we showed that this background torque is
negative for {\bf non-axisymmetric} trailing waves in discs with a fixed temperature or
sound-speed profile that decrease outwards. {\bf The background torque} enforces  
low-frequency {\bf non-axisymmetric} modes because they are associated
with negative angular momentum.

{\bf 
  We demonstrated the destabilising effect of the background torque by
  carrying out direct numerical hydrodynamic simulations of 
  locally isothermal discs with a self-gravitating surface density
  bump. 
}
We find such systems {\bf are} unstable to low-frequency perturbations
with azimuthal wavenumber $m=1$, which leads to the development of an one-armed 
trailing spiral that persist for at least $O(10^2)$ orbits. The spiral
pattern speed is smaller than the local disc rotation and
growth rates are $O(10^{-2}\Omega)$ which gives a characteristic
growth time of $O(10)$ orbits.

We used three independent numerical codes --- FARGO in 2D, ZEUS-MP and
PLUTO in 3D --- to show that the growth of     
one-armed spirals in our disc model is due to the imposed
temperature {\bf gradient: growth rates increased linearly}  
with the magnitude of the imposed temperature gradient, and one-armed
spirals did not develop in strictly isothermal simulations. {\bf This 
  one-armed spiral instability can be interpreted as an initially 
  neutral, tightly-wound $m=1$ mode being destabilised by the 
  background torque.} The spiral
is mostly confined between two $Q$-barriers surface density bump. 
We find the instability behaves similarly in 2D and 3D {\bf, but in 3D}
the spiral disturbance becomes more radially global away from    
the midplane.

\subsection{Speculations and future work}
% numerical simulations to include cooling/heating
% A locally isothermal equation of state represents the limit of
% infinitely short cooling (and heating) timescales, so the disc
% temperature instantly returns to its initial value when perturbed.

{\bf There are several issues that remain to be addressed in future
  works:}

{\bf \emph{Thermal relaxation.} The locally isothermal assumption 
} can be relaxed by including an energy equation with
{\bf a source term that restores the disc temperature over a
  characteristic timescale $t_\mathrm{relax}$}. Preliminary FARGO
simulations indicate a thermal relaxation timescale $t_\mathrm{relax} <
0.1\Omega_k^{-1}$ is needed for the one-armed spiral to
develop. However, this value may {\bf be model-dependent} and will be
explored in a follow-up study. {\bf For example,
a longer $t_\mathrm{relax}$ may be permitted with larger temperature
gradients.}

% prelim sims show they are long lasting 
% balancing heating and cooling 
% should not get fragmentation (CR away from spiral)

{\bf \emph{Non-linear evolution}.} In the deeply non-linear regime, the
one-armed spiral may  
shock and deposit negative angular momentum onto 
the background disc. The spiral amplitude would saturate by gaining
positive angular momentum. However, if the temperature gradient is
maintained, it may be possible to achieve a balance between the gain
of negative angular momentum through the background torque, and the
gain of positive angular momentum through shock dissipation. We remark  
that fragmentation {\bf of a low-frequency spiral} is unlikely
because the co-rotation radius is outside the bulk of the spiral arm, in the non-self-gravitating
portion of the disc \citep{durisen08}. In order
to study these possibilities, improved numerical models are needed to
ensure total angular momentum conservation on timescales much longer
than that considered in this paper.

@@

{\bf The background torque in other contexts.}  
{\bf In our simulations}, self-gravity coupled with the disc 
structure permits a neutral one-armed mode, which is then
destabilised by the {\bf fixed} temperature gradient.  
Other origins of the neutral one-armed spiral should be
investigated. One possibility is in Be star discs \citep{rivinius13},
for which one-armed oscillations may explain long-timescale variations
in their emission lines \citep[see e.g.][and references 
therein]{okasaki97,papaloizou06c,ogilvie08}. These studies employ
strictly isothermal disc models, so it would be interesting to explore
the effect of a temperature gradient on the stability of these
oscillations.

%vertical shear instability?
%linear theory: explicit calculations 
%For a self-consistent description of the instability, one needs to
%explicitly solve for linear stability problem. 
%theoretical investigation on effect on other m modes
